{\bf Intrusion Detection through Provenance-Based Histograms}

### 2013-11-24 writeup/paper.tex
Expanded methodology. 2 more figures

\section{Design and Implementation}
Here we present our intrusion detection technique: a general and extendable statistical method for analyzing and comparing provenance data. At a high level, we process a provenance data graph, generating a model of "normal" behavior for each process/file in the graph. Using this model, we can determine the likelihood that a given "test" node is normal or whether it deviates from normal behavior and is thus potentially an exploited or malicious program.
\subsection{Models}
Our approach requires statistical models to evaluate the normalcy of a test node. The statistical model we chose to use was a kernel density estimation (KDE), which estimates the probability density function of a value with unknown distribution. To create the KDE, we needed to provide counts of a particular statistic. In this way, for each statistic that we chose to collect, we would be able to evaluate a test node's normalcy according to that statistic. For example, we could then ask "Does node $X_n$ have a statistically believable number of output processes compared to the other $X$ nodes?"
\subsection{Provenance Statistics}
As described above, in order to generate a KDE model, we need some statistic to measure. As we expected intrusions to differ in terms of the statistics that would stand out, we needed to generate a large number of statistics, and a model for each. With regards to provenance nodes directly, we compiled the following statistics:
\begin{enumerate}
\item The number of input files to the node.
\item The number of input processes to the node.
\item The number of input pipes to the node.
\item The number of output files to the node.
\item The number of output processes to the node.
\item The number of ouput pipes to the node.
\item file\_in\_version ??? XXX
\item file\_out\_version ??? XXX
\end{enumerate}
For ease of computation and comparison, we treated the different node types (e.g. file, process, pipe) separately so that we actually had different counts for "input files to a file node" and "input files to a process node", for instance.
\subsection{Centrality Metrics}
In addition to the data that can be collected on the nodes directly, we could also compute clustering statistics by considering each node in the context of the larger graph. Using the methods espoused by Macko et al. \cite{clustering}, we could compute different centrality and clustering metrics for a node, which can then be tabulated into a histogram like the input/output counts above. 
KENNY TODO: EXPAND THIS
\subsection{Implementation}
The implementation of the above techniques consists of multiple Python files, spanning about 850 lines of (commented) code. The raw BerkeleyDB database generated by PASS is provided as input, and textual/graphical results are outputted.
KENNY TODO: EXPAND THIS
\subsection{Examples}
In Figure 1, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a "test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \caption{Example histograms of a centrality measure and process statistic, with the assoicated KDE}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}

@@

\subsection{Evaluation Methodology}
In order to test our techniques, we needed to run “exploits” on our PASS-enabled machine. We looked to prior work to get a sense of the different types of techniques commonly used to exploit security vulnerabilities. Commonly, remote network techniques (as seen in Backtracker \cite{backtracker}) are used to broach insecure systems. Metasploit [CITE] is a software package designed to facilitate running exploits on insecure system. Searching the Metasploit database, as well as the online Exploit Database (exploit-db.com) [CITE], we found that most successful techniques involve exploiting vulnerabilities in software that a target system is running, as opposed to vulnerabilities in the target system itself. Put another way, we sought to exploit software packages, like mcrypt and gcc, not the Linux kernel itself.

@@

\subsection{Results}
\subsection{Provenance Data}
After running the normal and exploited programs described above, we visualized the provenance graphs using Orbiter \cite{orbiter}. After searching for the nodes named after our programs, we were able to find both the clean and exploited nodes. The clean nodes for "hello" simply read in one file through a pipe and outputted one file, whereas the exploited version clearly has a forked "/usr/bin/ls" process. Similarly, the mcrypt simulation (implemented in Python) has only an output file as a descendent normally, but when exploited has multiple extra processes/files as descendents (see Figure 2). Similarly, the normal GCC and exploited GCC displayed very different provenance graphs. 
\begin{figure*}
  \caption{Descendents of an unexploited mcrypt-style program. Simply an output file. Below, descendents of an exploited mcrypt-style program. In addition to an (offscreen) output file, there is also an outgoing /bin/sh process, with its own children.}
  \centering
    \includegraphics[width=\textwidth]{img/mcrypt.png}
\end{figure*}

\subsection{Statistical Data and KDE Predictions}
Table 1 summarizes some of the preliminary results we have generated. Here we have included, for each statistic and test, the average over the "normal" nodes, the observed value for the "test" node, and the KDE probability output for that value. We have chosen four metrics that provided a statistically significant result. The other statistics failed to be as powerful (see Table 2 XXX).
% from http://hhh123.wordpress.com/2008/08/13/latex-table-across-two-columns/
\begin{center}
\begin{table*}[ht]
{\small
  \begin{tabular}{| l | l | l | l |}
    \hline
    metric & hello & mcrypt & gcc \\ \hline
     \# forked processes & 0.0 | 1 | 0.108 & 0.0 | 1 | 0.107 & 0.9 | 9 | 0.408 \\ \hline
     \# input files & 0.5 | 0 | 0.453 & 0.5 | 0 | 0.453 & 0.5 | 1 | 0.107 \\ \hline
     \# ouptut files & 0.5 | 1 | 0.453 & 0.5 | 1 | 0.453 & 0.0 | 3 | 0.000 \\ \hline
    Opsahl centrality & 1.791 | 2.83 | 0.252 & 1.250 | 4.867 | 0.000 & 2.793 | 7.415 | 0.024 \\
    \hline
  \end{tabular}
}
\hfill{}
\caption{average over good | test node value | kde(test node is normal)
}
\label{tb:tablename}
\end{table*}
\end{center}

@@

\end{thebibliography}

>>>>

\section{Design and Implementation}
Here we present our intrusion detection technique: a general and extendable statistical method for analyzing and comparing provenance data. At a high level, we process a provenance data graph, generating a model of "normal" behavior for each process/file in the graph. Using this model, we can determine the likelihood that a given "test" node is normal or whether it deviates from normal behavior and is thus potentially an exploited or malicious program.
\subsection{Models}
Our approach requires statistical models to evaluate the normalcy of a test node. The statistical model we chose to use was a kernel density estimation (KDE), which estimates the probability density function of a value with unknown distribution. To create the KDE, we needed to provide counts of a particular statistic. In this way, for each statistic that we chose to collect, we would be able to evaluate a test node's normalcy according to that statistic. For example, we could then ask "Does node $X_n$ have a statistically believable number of output processes compared to the other $X$ nodes?"
\subsection{Provenance Statistics}
As described above, in order to generate a KDE model, we need some statistic to measure. As we expected intrusions to differ in terms of the statistics that would stand out, we needed to generate a large number of statistics, and a model for each. With regards to provenance nodes directly, we compiled the following statistics:
\begin{enumerate}
\item The number of input files to the node.
\item The number of input processes to the node.
\item The number of input pipes to the node.
\item The number of output files to the node.
\item The number of output processes to the node.
\item The number of ouput pipes to the node.
\item file\_in\_version The "version" number of an incoming file.
\item file\_out\_version The "version" number of an outgoing file.
\end{enumerate}
For ease of computation and comparison, we treated the different node types (e.g. file, process, pipe) separately so that we actually had different counts for "input files to a file node" and "input files to a process node", for instance.
\subsection{Centrality Metrics}
In addition to the data that can be collected on the nodes directly, we could also compute clustering statistics by considering each node in the context of the larger graph. Using the methods espoused by Macko et al. \cite{clustering}, we could compute different centrality and clustering metrics for a node, which can then be tabulated into a histogram like the input/output counts above. 
KENNY TODO: EXPAND THIS
\subsection{Implementation}
The implementation of the above techniques consists of multiple Python files, spanning about 850 lines of (commented) code. The raw BerkeleyDB database generated by PASS is provided as input, and textual/graphical results are outputted.
KENNY TODO: EXPAND THIS
\subsection{Examples}
In Figure 1, we see a few examples of a generated histogram and KDE model. The green bars represent the histogram for the particular statistic, and the blue curve is the KDE. Finally, the red line is the value for a "test" node. In this case, the test node appears to be a possible exploit, because it intersects the KDE at low values. 
\begin{figure*}
  \caption{Example histograms of a centrality measure and process statistic, with the assoicated KDE}
  \centering
    \includegraphics[width=\textwidth]{img/hist.png}
\end{figure*}

@@

\subsection{Evaluation Methodology}
In order to test our techniques, we needed to run “exploits” on our PASS-enabled machine. We looked to prior work to get a sense of the different types of techniques commonly used to exploit security vulnerabilities. Commonly, remote network techniques (as seen in Backtracker \cite{backtracker}) are used to broach insecure systems. Metasploit \cite{metasploit} is a software package designed to facilitate running exploits on insecure system. Searching the Metasploit database, as well as the online Exploit Database \cite{exploitdb}, we found that most successful techniques involve exploiting vulnerabilities in software that a target system is running, as opposed to vulnerabilities in the target system itself. Put another way, we sought to exploit software packages, like mcrypt and gcc, not the Linux kernel itself.

@@

\subsection{Selecting Metrics}
Because we had no prior intuition into what the provenance data would actually look like, we approached selecting metrics from an objective, data-based standpoint. We wanted to use metrics that had variation over our workload, so as not to minimize the introduction of workload dependent data. Towards this end, we graphed all of the metrics over the "normal" provenance data, to see where there were trends and variation. Our results for GCC are in Figures 2 and 3. We can see that proc\_out\_proc (outgoing processes) and the ancestor and opsahl centrality metrics have more variation, and will be less likely to introduce biases. As we include more data (as discussed in Future Works), these sorts of graphs will become increasingly important in selecting good metrics for evaluation.
\begin{figure}
  \caption{Histograms over all "simple" provenance statistics for GCC}
  \centering
    \includegraphics[width=0.5\textwidth]{img/gcc_stats.png}
\end{figure}
\begin{figure}
  \caption{Histograms over all centrality metrics for GCC}
  \centering
    \includegraphics[width=0.5\textwidth]{img/gcc_cent_stats.png}
\end{figure}

\subsection{Results}
\subsection{Provenance Data}
After running the normal and exploited programs described above, we visualized the provenance graphs using Orbiter \cite{orbiter}. After searching for the nodes named after our programs, we were able to find both the clean and exploited nodes. The clean nodes for "hello" simply read in one file through a pipe and outputted one file, whereas the exploited version clearly has a forked "/usr/bin/ls" process. Similarly, the mcrypt simulation (implemented in Python) has only an output file as a descendent normally, but when exploited has multiple extra processes/files as descendents (see Figure 4). Similarly, the normal GCC and exploited GCC displayed very different provenance graphs. 
\begin{figure*}
  \caption{Descendents of an unexploited mcrypt-style program. Simply an output file. Below, descendents of an exploited mcrypt-style program. In addition to an (offscreen) output file, there is also an outgoing /bin/sh process, with its own children.}
  \centering
    \includegraphics[width=\textwidth]{img/mcrypt.png}
\end{figure*}

\subsection{Statistical Data and KDE Predictions}
Table 1 summarizes some of the preliminary results we have generated. Here we have included, for each statistic and test, the average over the "normal" nodes, the observed value for the "test" node, and the KDE probability output for that value. We have chosen four metrics that provided a statistically significant result.
% The other statistics failed to be as powerful (see Table 2 XXX).
% from http://hhh123.wordpress.com/2008/08/13/latex-table-across-two-columns/
\begin{center}
\begin{table*}[ht]
{\small
  \begin{tabular}{| l | l | l | l |}
    \hline
    metric & hello & mcrypt & gcc \\ \hline
     \# forked processes & 0.0 | 1 | 0.108 & 0.0 | 1 | 0.107 & 0.9 | 9 | 0.408 \\ \hline
     \# input files & 0.5 | 0 | 0.453 & 0.5 | 0 | 0.453 & 0.5 | 1 | 0.107 \\ \hline
     \# ouptut files & 0.5 | 1 | 0.453 & 0.5 | 1 | 0.453 & 0.0 | 3 | 0.000 \\ \hline
    Opsahl centrality & 1.791 | 2.83 | 0.252 & 1.250 | 4.867 | 0.000 & 2.793 | 7.415 | 0.024 \\
    \hline
  \end{tabular}
}
\hfill{}
\caption{average over good | test node value | kde(test node is normal)
}
\label{tb:tablename}
\end{table*}
\end{center}

@@

\bibitem{metasploit}
\textsc{Rapid 7 Inc.} Metasploit Framework. \tt{http://www.metasploit.com}.

\bibitem{exploitdb}
\textsc{Offensive Security, Inc.} The Exploit Database. \tt{http://www.exploit-db.com}.
\end{thebibliography}
